Supplementary Materials to ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic''

### 2022-09-19 write-up/covid-supplement-writeup.tex
updating sensitivity model and making all relevant changes for final referee comments
Sensitivity analysis complete :)

\section{Sensitivity Analysis}

The proposed approach relies on the assumption $I_{j,t}^{(NR)} \indep Y_{j,t} \, | \, X_{j,t}$, i.e., the sampling indicator in the non-probabilistic survey is conditionally independent of the outcome given the covariates.  However, estimates of population quantities based on inverse-weighting and/or model-based estimates will be biased in the presence of `unobserved confounding'.  Here we introduce a sensitivity analysis to address this concern.  To this end, we consider the logit-linear model by~\cite{Imbens2003}, adjusted to the time-varying, self-selection setting from the single time-point, causal inference setting.  That is, consider the model
\begin{align*}
\logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) &= h_t(x_t) + \alpha_t u_t \\
E[ Y_t | i_t, x_t, u_t] &= l_t(i_t, x_t) + \delta_t u_t
\end{align*}
for functions~$h_t$ and $l_t$ that depend on time. Re-arranging terms we have
$$
E[ Y_t | i_t x_t, u_t ] = l_t(i_t,x)  + \frac{\delta_t}{\alpha_t} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
$$
The key idea in~\cite{Imbens2003} is that positing a distribution on $\pi(I_t = 1 | x_y, u_t)$ directly allows one to circumvent the need to specify a distribution for~$U_t$, a highly non-trivial task.

The logit-linear model above unfortunately does not lead to a tractable sensitivity analysis.  Instead, we consider a sensitivity model based on work by~\cite{Veitch2020}, again adjusted to deal with the current setting of self-selection, given by:
\begin{equation}
\begin{aligned}
\label{eq:sensmodel}
\pi_t(X_t, U_t) &\sim \text{Beta} \left( \pi_t (X_t) (1/\alpha_t - 1), (1-\pi_t(X_t)) (1/\alpha_t - 1) \right) \\
I_t | X_t, U_t &\sim \text{Bern} (\pi (X_t,U_t)) \\
\rho_t(X_t, U_t) &= Q_t(X_t,1) + \delta_t \left( \logit \pi_t(X_t,U_t) - \EE \left[ \logit \pi_t(X_t,U_t) | X_t, I_t = 1 \right] \right) \\
Y_t | I_t, X_t, U_t &\sim \text{Bern}(\rho_t(X_t, U_t))
\end{aligned}
\end{equation}
The time-varying sensitivity parameter~$\alpha_t \in (0,1)$  controls the influence of the unobserved confounder~$U_t$ on selection propensity. In particular, $\alpha_t$ measures the change in belief of how likely an individual self-selects into the non-probabilistic sample at time~$t$:
$$
\alpha_t = E[ \pi_t (X_t,U_t) | I = 1] - E[ \pi_t (X_t,U_t) | I = 0].
$$
The sensitivity model satisfies the requirement that the self-selection propensity and conditional expectation outcome model can match the observed data:
\begin{align*}
P(I_t = 1 | X_t ) &= E[ E[ I_t | X_t, U_t ] | X_t ] = E[ \pi_t(X_t, U_t) | X_t ] = \pi_t (X_t) \\
E[Y_t | X_t, I_t = 1 ] &= E[ E[ Y_t | X_t, U_t, I_t =1 ] | X_t, I_t = 1 ] =: Q_t(1,X_t).
\end{align*}
Moreover, the model satisfies that $Y_t$ does not depend on $I_t$ given $(X_t, U_t)$. This is a key adaptation to the non-probabilistic survey setting with potential time-varying confounding and a binary outcome.

@@

Following~\cite{Veitch2020}, we re-express the outcome-confounder strength in terms of the partial coefficient of determination
$$
R_t^2 (\alpha_t, \delta_t) = \frac{\EE [( Y_{J,t} - Q_t(1,X_{J,t}) )^2 | I_{J,t}=1 ] - \EE[ (Y_{J,t} - \EE[ Y_{J,t} | X_{J,t}, U_{J,t} ])^2 | I_{J,t} = 1]}{E[(Y_{J,t} - Q_t(1,X_{J,t}))^2 | I = 1]}
$$
in terms of $\delta_t^2$. To do so, write
\begin{align*}
&\EE [ (Y_{J,t} - \EE[ Y_{J,t} | X_{J,t}, U_{J,t}] )^2 | I_{J,t} = 1 ] \\
= &\EE [ (Y_{J,t} - Q_t(1,X_{J,t}))^2 | I_{J,t} = 1 ] \\
&- 2 \delta \EE \left[ (Y_{J,t} - Q_t(1,X_{J,t})) (\logit \pi(X_{J,t},U_{J,t}) - \EE[ \logit \pi (X_{J,t},U_{J,t}) | X_{J,t}, I_{J,t}=1 ]) | I_{J,t} = 1 \right] \\
&+ \delta^2 \EE \left[ (\logit \pi_t(X_{J,t},U_{J,t}) - \EE[ \logit \pi_t (X_{J,t},U_{J,t}) | X_{J,t}, I_{J,t}=1 ])^2 | I_{J,t} = 1 \right] \\
= &\EE [ (Y_{J,t} - Q_t(1,X_{J,t}))^2 | I_{J,t} = 1 ] - \delta^2 \EE \left[ \text{var} \left( \logit \pi_t(X_{J,t},U_{J,t}) \right) | X_{J,t}, I_{J,t} = 1 \right]
\end{align*}
By Beta-Bernoulli conjugacy, the second term is the variance of the logit-transformed Beta distribution which has an analytic expression:
$$
\text{var} \left( \logit \pi_t(X_{J,t},U_{J,t}) | X_{J,t}, I_{J,t} = 1 \right) = \psi_1 ( \pi(X_{J,t}) (\alpha_t - 1) + 1 ) + \psi_1 ( (1- \pi(X_{J,t})) (1/\alpha_t - 1))
$$
where $\psi_1$ is the trigamma function. This implies the following relationship:
\begin{equation}
\label{eq:partialcoef}
R_t^2 (\alpha_t, \delta_t) = \delta_t^2 \frac{\EE [\psi_1 ( \pi(X_{J,t}) (\alpha_t - 1) + 1 ) + \psi_1 ( (1- \pi(X_{J,t})) (1/\alpha_t - 1)) | I_{J,t} = 1 ] }{E [(Y_{J,t}- Q_t(1,X_{J,t}))^2 | I_{J,t} = 1]}
\end{equation}
Note that unlike the prior expecation, these are expressed conditional on self-selection, i.e., $I_{J,t} = 1$.  This was done since the binary outcomes are only measured in the non-probabilistic sample.

@@

Based on~\eqref{eq:partialcoef} and~\eqref{eq:alphaest}, one can use the probability and nonprobability samples to calibrate the sensitivity parameters to be in line with the observed covariate influence on treatment and outcome.  That is, for a given observed covariate~$Z_{J,t}$, we wish to measure the degree of influence it has on self-selection and outcome given the other observed covariates~$X_{J,t} \backslash Z_{J,t}$.  Note that this dependence is likely time-varying, which is why it is important estimate~$\alpha_t$ and~$R_t^2$ as functions of time~$t$.

For the outcome, we can measure the partial coefficient of determination as:
$$
R_{t, Z}^2 := \frac{\frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t} \backslash z_{i,t}) )^2 - \frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t}) )^2}{\frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t} \backslash z_{i,t}) )^2}.
$$
We can measure influence of $Z$ on self-selection propensity given $X_{J,t} \backslash Z_{J,t}$ by
$$
\hat \alpha_{t, Z} := 1 - \frac{\sum_{j=1}^N I_{j,t}^R W_{j,t}^R \cdot \hat \pi(X_{J,t}) (1- \hat \pi(X_{J,t}))}{\sum_{j=1}^N I_{j,t}^R W_{j,t}^R \cdot \hat \pi(X_{J,t} \backslash Z_{J,t}) (1- \hat \pi(X_{J,t} \backslash Z_{J,t}))}
$$
where~$\hat \pi_t (\cdot)$ is the fitted self-selection probability based on~\eqref{eq:auxinfoprob} in Section~\ref{subsec:auxprob} using the appropriately chosen set of observed covariates.

@@

Here we perform a sensitivity analysis of the IPW estimator of COVID-19 active infection prevalence in indiana.  While~\cite{Veitch2020} use Austen plots, we do not have such a simple visualization due to time-dependence.  To handle this, for each observed covariate~$Z$, we plot $(\alpha_{t,Z}, R^2_{t,Z})$ in the $[0,1]^2$.  These ``calibration curves'' help us in selection of a range of values for~$(\alpha_t, R^2_t)$ in our sensitivity analysis. Here, we consider $Z$ to be separately (a) age, (b) gender, (c) race, and (d) symptom status.  The four plots are presented in Figure XX.

>>>>

\section{Sensitivity Analysis to Unmeasured Confounding}
\label{sec:sensitivity_confounders}

The proposed approach relies on the assumption $I_{j,t}^{(NR)} \indep Y_{j,t} \, | \, X_{j,t}$, i.e., the sampling indicator in the non-probabilistic survey is conditionally independent of the outcome given the covariates.  However, estimates of population quantities based on inverse-weighting and/or model-based estimates will be biased in the presence of `unobserved confounding'.  Here we introduce a sensitivity analysis to address this concern.  To this end, we consider the logit-linear model by~\cite{Imbens2003}, adjusted to the time-varying, self-selection setting from the single time-point, causal inference setting.  That is, consider the model
\begin{align*}
\logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) &= h_t(x_t) + \alpha_t u_t \\
E[ Y_t | i_t, x_t, u_t] &= l_t(i_t, x_t) + \delta_t u_t
\end{align*}
for functions~$h_t$ and $l_t$ that depend on time. Re-arranging terms we have
$$
E[ Y_t | i_t, x_t, u_t ] = l_t(i_t,x)  + \frac{\delta_t}{\alpha_t} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
$$
The key idea in~\cite{Imbens2003} is that positing a distribution on $\pi(I_t = 1 | x_y, u_t)$ directly allows one to circumvent the need to specify a distribution for~$U_t$, a highly non-trivial task.

The logit-linear model above unfortunately does not lead to a tractable sensitivity analysis.  Instead, we consider a sensitivity model based on work by~\cite{Veitch2020}, again adjusted to deal with the current setting of self-selection, given by:
\begin{equation}
\begin{aligned}
\label{eq:sensmodel}
\pi_t(X_t, U_t) &\sim \text{Beta} \left( \pi_t (X_t) (1/\alpha_t - 1), (1-\pi_t(X_t)) (1/\alpha_t - 1) \right) \\
I_t | X_t, U_t &\sim \text{Bern} (\pi (X_t,U_t)) \\
\rho_t(X_t, U_t) &= Q_t(X_t,1) + \delta_t \left( \logit \pi_t(X_t,U_t) - \EE \left[ \logit \pi_t(X_t,U_t) | X_t, I_t = 1 \right] \right) \\
Y_t | I_t, X_t, U_t &\sim \text{Bern}(\rho_t(X_t, U_t))
\end{aligned}
\end{equation}
The time-varying sensitivity parameter~$\alpha_t \in (0,1)$  controls the influence of the unobserved confounder~$U_t$ on selection propensity. In particular, $\alpha_t$ measures the change in belief of how likely an individual self-selects into the non-probabilistic sample at time~$t$:
$$
\alpha_t = E[ \pi_t (X_t,U_t) | I_t = 1] - E[ \pi_t (X_t,U_t) | I_t = 0].
$$
The sensitivity model satisfies the requirement that the self-selection propensity and conditional expectation outcome model can match the observed data:
\begin{align*}
P(I_t = 1 | X_t ) &= E[ E[ I_t | X_t, U_t ] | X_t ] = E[ \pi_t(X_t, U_t) | X_t ] = \pi_t (X_t) \\
E[Y_t | X_t, I_t = 1 ] &= E[ E[ Y_t | X_t, U_t, I_t =1 ] | X_t, I_t = 1 ] =: Q_t(1,X_t).
\end{align*}
Moreover, the model satisfies that $Y_t$ does not depend on $I_t$ given $(X_t, U_t)$. This is a key adaptation to the non-probabilistic survey setting with potential time-varying confounding and a binary outcome.

@@

Following~\cite{Veitch2020}, we re-express the outcome-confounder strength in terms of the partial coefficient of determination
$$
R_t^2 (\alpha_t, \delta_t) = \frac{\EE [( Y_{J,t} - Q_t(1,X_{J,t}) )^2 | I_{J,t}=1 ] - \EE[ (Y_{J,t} - \EE[ Y_{J,t} | X_{J,t}, U_{J,t} ])^2 | I_{J,t} = 1]}{E[(Y_{J,t} - Q_t(1,X_{J,t}))^2 | I_t = 1]}
$$
in terms of $\delta_t^2$. To do so, write
\begin{align*}
&\EE [ (Y_{J,t} - \EE[ Y_{J,t} | X_{J,t}, U_{J,t}] )^2 | I_{J,t} = 1 ] \\
= &\EE [ (Y_{J,t} - Q_t(1,X_{J,t}))^2 | I_{J,t} = 1 ] \\
&- 2 \delta \EE \left[ (Y_{J,t} - Q_t(1,X_{J,t})) (\logit \pi(X_{J,t},U_{J,t}) - \EE[ \logit \pi (X_{J,t},U_{J,t}) | X_{J,t}, I_{J,t}=1 ]) | I_{J,t} = 1 \right] \\
&+ \delta^2 \EE \left[ (\logit \pi_t(X_{J,t},U_{J,t}) - \EE[ \logit \pi_t (X_{J,t},U_{J,t}) | X_{J,t}, I_{J,t}=1 ])^2 | I_{J,t} = 1 \right] \\
= &\EE [ (Y_{J,t} - Q_t(1,X_{J,t}))^2 | I_{J,t} = 1 ] - \delta^2 \EE \left[ \text{var} \left( \logit \pi_t(X_{J,t},U_{J,t}) \right) | X_{J,t}, I_{J,t} = 1 \right]
\end{align*}
By Beta-Bernoulli conjugacy, the second term is the variance of the logit-transformed Beta distribution which has an analytic expression:
$$
\text{var} \left( \logit \pi_t(X_{J,t},U_{J,t}) | X_{J,t}, I_{J,t} = 1 \right) = \psi_1 ( \pi(X_{J,t}) (\alpha_t - 1) + 1 ) + \psi_1 ( (1- \pi(X_{J,t})) (1/\alpha_t - 1))
$$
where $\psi_1$ is the trigamma function. This implies the following relationship:
\begin{equation}
\label{eq:partialcoef}
R_t^2 (\alpha_t, \delta_t) = \delta_t^2 \frac{\EE [\psi_1 ( \pi(X_{J,t}) (\alpha_t - 1) + 1 ) + \psi_1 ( (1- \pi(X_{J,t})) (1/\alpha_t - 1)) | I_{J,t} = 1 ] }{E [(Y_{J,t}- Q_t(1,X_{J,t}))^2 | I_{J,t} = 1]}
\end{equation}
Note that unlike the prior expecation, these are expressed conditional on self-selection, i.e., $I_{J,t} = 1$.  This was done since the binary outcomes are only measured in the non-probabilistic sample.

@@

Based on~\eqref{eq:partialcoef} and~\eqref{eq:alphaest}, one can use the probability and nonprobability samples to calibrate the sensitivity parameters to be in line with the observed covariate influence on treatment and outcome.  That is, for a given observed covariate~$Z_{J,t}$, we wish to measure the degree of influence it has on self-selection and outcome given the other observed covariates~$X_{J,t} \backslash Z_{J,t}$.  Note that this dependence is likely time-varying, which is why it is important that~$\alpha_t$ and~$R_t^2$ are functions of time~$t$.

For the outcome, we can measure the partial coefficient of determination as:
$$
R_{t, Z}^2 := \frac{\frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t} \backslash z_{i,t}) )^2 - \frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t}) )^2}{\frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t} \backslash z_{i,t}) )^2}.
$$
among the non-probabilistic sample.  We can measure influence of $Z$ on self-selection propensity given $X_{J,t} \backslash Z_{J,t}$ by
$$
\hat \alpha_{t, Z} := 1 - \frac{\sum_{j=1}^N I_{j,t}^R W_{j,t}^R \cdot \hat \pi(X_{J,t}) (1- \hat \pi(X_{J,t}))}{\sum_{j=1}^N I_{j,t}^R W_{j,t}^R \cdot \hat \pi(X_{J,t} \backslash Z_{J,t}) (1- \hat \pi(X_{J,t} \backslash Z_{J,t}))}
$$
where~$\hat \pi_t (\cdot)$ is the fitted self-selection probability based on~\eqref{eq:auxinfoprob} in Section~\ref{subsec:auxprob} using the appropriately chosen set of observed covariates.

@@

Here we perform a sensitivity analysis of the IPW estimator of COVID-19 active infection prevalence in indiana.  While~\cite{Veitch2020} use Austen plots, we do not have such a simple visualization due to time-dependence.  Here, we plot  $\alpha_{t,Z}$ and $R^2_{t,Z}$ as functions of time for three covariates that demonstrate different patterns: (1) Symptoms, (2) Race, and (3) Ethnicity.  Figures~\ref{fig:alpha_sensitivity} and~\ref{fig:rsq_sensitivity} present these ``calibration curves.'' Figure~\ref{fig:bias_sensitivity} presents the bias under these three curves.  Here, we see that if an unobserved confounder has similar strength in selection-confounder and outcome-confounder relationships as ``Symptom Status'', then bias would be high even into 2022.  However, bias under an unobserved confounder with similar strength to ``Ethnicitiy'' dissipates significantly by late 2021.

\begin{figure}[!th]
 \centering
 \begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/alpha_sensitivity_fig.png}
  \caption{Calibration curves for~$\alpha$ per week}
  \label{fig:alpha_sensitivity}
 \end{subfigure}
 \begin{subfigure}{.45\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/rsq_sensitivity_fig.png}
 \caption{Calibration curves for~$R^2$ per week}
 \label{fig:rsq_sensitivity}
 \end{subfigure} \\ [1ex]
 \begin{subfigure}{\linewidth}
 \centering
 \includegraphics[width=.5\linewidth]{../figs/bias_sensitivity_fig.png}
  \caption{Bias as a function of~$\alpha$ and~$R^2$ calibration curves}
  \label{fig:bias_sensitivity}
 \end{subfigure}
 \caption{Sensitivity analysis for time-varying active infection rate estimates}
 \label{fig:sensitivity}
 \end{figure}
